Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-09-18 inst/doc/nonparametric-bayes.Rmd
writing writing

@@

<!--
While methods that account for _parametric_ uncertainty can be very 
successful with the right model,
structural uncertainty of not knowing what model best approximates the 
dynamics poses considerably greater difficulty.  
-->

@@

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.  
Here we illustrate how a stochastic dynamic programming (SDP) algorithm can 
be driven by the predictions from a Gaussian Process (GP).
This provides two distinct advantages compared with contemporary approaches.
First, using a GP sidesteps the need for an accurate model-based description
of the system dynamics (or a model choice or model-averaging approach which
introduces yet more difficulty).  
Second, unlike parametric model-fitting approaches, which can only reflect 
uncertainty in the corresponding confidence intervals or  posterior distributions
of the parameters, the GP approach provides a state-space dependent representation
of uncertainty.  In other words, a GP estimated from a given data set permits 
a much greater uncertainty far from the observed data than it does near the observed 
data.

@@

Cite [@Glaser2013](http://doi.org/10.1111/faf.12037 "Complex dynamics may limit prediction in marine fisheries")
etc

also older stuff on model uncertainty

@@

<!-- More on the lack of data throughout the relevant state-space
     and how any concern about potential tipping points indicates 
     that the data exhibits this bias / problem.

The idea that any approach can perform well without either having to know the model or
have particularly good data should immediately draw suspicion.  The reader must bear
in mind that the strength of our approach comes not from some magic predictive power
from such limited information, but rather, by providing a more honest expression of 
uncertainty outside the observed data without sacrificing the predictive capacity 
near the observed data.

By coupling this more accurate description of what is known and unknown to the 
decision-making under uncertainty framework provided by stochastic dynamic programming,
we are able to obtain more robust management policies than with common parameteric modeling 
approaches.

@@

Unfortunately, estimates of parameter uncertainty are only as good as the parametric 
models themselves.  Too often, we do not understand the system dynamics well enough
to know if a model provides a good approximation over the relevant range of states
and timescales (criteria that we loosely refer to as defining the "right" or "true" model.)
So called structural or model uncertainty is a more difficult problem than parametric 
uncertainty.  Typical solutions involve either model choice, or model averaging. Either
approach remains limited by the degree to which any of the proposed models are sufficiently
close to the right model.

@@

#### Comparing models

@@

#### Background on Fisheries Context

#### Background on Tipping points

<!-- note: This section largely replaced by above outline

### Quantitative vs Qualitative Decisions

@@

Non-parametric models can better represent uncertainties outside 
of observed data while also better capturing the dynamics in the region
observed.  Advances in the theory and computational implementations of 
nonparametric methods make them ripe for such applications.  We use the classic
problem of optimal harvest of a marine fishery to illustrate how the 
nonparametric approach of Gaussian processes can be applied.  We will 
compare Bayesian implementations of both nonparametric and parametric 
models, which best allow us to capture the uncertainty of model estimation
in either case and permits a more natural comparison between approaches.

@@

In our example, we focus on the problem in which a manager must set 
the harvest level for a marine fishery each year to maximize the net
present value of the resource, given an estimated stock size from the year
before. The economic value and ecological concern have made marine fisheries the crucible for much
of the founding work [@Gordon1954; @Reed1979; @May1979; @Ludwig1982]
in managing ecosystems under uncertainty.  Global trends [@Worm2006]
and controversy [@Hilborn2007; @Worm2009] have made understanding these
challenges all the more pressing.

>>>>

Abstract
========

Model uncertainty and limited data coverage are fundamental challenges to
robust ecosystem management.  These challenges are acutely highlighted
by concerns that many ecological systems may contain tipping points.
Before a collapse, we do not know where the tipping points lie, if the
exist at all.  Hence, we know neither a complete model of the system
dynamics nor do we have access to data in some large region of state-space
where such a tipping point might exist.  These two sources of uncertainty
frustrate state-of-the-art parametric approaches to decision theory
and optimal control.  I will illustrate how a non-parametric approach
using a Gaussian Process prior provides a more flexible representation
of this inherent uncertainty.  Consequently, we can adapt the Gaussian
Process prior to a stochastic dynamic programming framework in order to
make robust management predictions under both model and uncertainty and
limited data.

Introduction
============

<!--
Very briefly set the stage: (see related papers)
- Decision-theory / optimal control framework
- Fisheries context 
-->

The sudden collapse of fisheries and other ecosystems is an increasingly
widespread phenomenon and a pressing concern for ecological management
and conservation.  Ecological dyanaics are frequently complex and difficult to 
measure, making uncertainty in our understanding a prediction a persistent
challenge to effective management.

Decision-theoretic approaches provide a framework to determine the best 
sequence of actions in face of uncertainty, but only when that uncertainty
can be meaningfully quantified [@Fischer2009].

Uncertainty enters at many levels <!-- enters what -->
management

- Why fisheries

The economic value and ecological concern have made marine
fisheries the crucible for much of the founding work [@Gordon1954;
@Reed1979; @May1979; @Ludwig1982] in managing ecosystems under
uncertainty.  Global trends [@Worm2006] and controversy [@Hilborn2007;
@Worm2009] have made understanding these challenges all the more pressing.

<!-- Other reasons for failure see Sethi-->

@Fischer 
@Sethi2005, @Peretti, @Munch,

(Classical division)

- Dynamic uncertainty (stochasticity)

From [Clark (1976)]() to [Reed (1979)]()

Other sources of uncertainty [Roughgarden & Smith 1996](), [Sethi _et al._ (2005)]()

- Parametric uncertainty <!-- ugh, be more consise and less annoying-->

[Ludwig & Walters 1982](http://doi.org/10.1016/0304-3800(82)90023-0 "Optimal harvesting with imprecise parameter estimates")

- Structural uncertainty <!-- ugh, be more consise and less annoying-->

Estimates of parameter uncertainty are only as good as the parametric
models themselves.  Often we do not understand the system dynamics well
enough to know if a model provides a good approximation over the relevant
range of states and timescales (criteria that we loosely refer to as
defining the "right" or "true" model.)  So called structural or model
uncertainty is a more difficult problem than parametric uncertainty.
Typical solutions involve either model choice, model averaging, or
introducing yet greater model complexity of which others may be special
cases (model averaging being one such way to construct such a model)
[@Williams2001; @Athanassoglou2012; @Cressie2009].  Even setting aside
other computational and statistical concerns (e.g. [@Cressie2009]), these
approaches do not address our second concern - representing uncertainty
outside the observed data range.

[@Cressie2009](http://dx.doi.org/10.1890/07-0744.1 "Accounting for uncertainty in ecological analysis: the strengths and limitations of hierarchical statistical modeling.").

- Limits of state space

Often the set of feasible states is larger than the range of the observed states.  Though a system can enter a state outside previously observed values due to stochastic effects alone (such as a particularly low or high nutrient year), this concern is most keenly felt in considering the value of actions that have not yet been applied.

Of all sources of uncertainty, this has potentially recieved the least attention.  It has most often appeared in the concept of active learning or active adaptive managment, (adaptive probing, ) where it may be valauble to intentionally force a system far from the observed values even when the expected value such actions is low, as it provides much faster learning and consequent reduction of model uncertainty that can allow greater value to be derived later on.  @Ludwig1982 show that it may be advantageous to fish an unexploited population very heavily at first to obtain a better estimate of the recruitment rate.  This intuitive strategy when a population is governed by a Ricker or Beverton-Holt-like dynamic would clearly be disasterous if instead the dynamics contained an unforseen tipping point.  The best way to learn where the edge lies may be to walk up to it, but it is also the most dangerous.

## Two central challenges:

- We don't know the model
- We don't have data from where we need it most

## Why we don't know the model

- Complex dynamics
- model choice and model averaging approaches

## Why we don't have data where we need it

- Concerns of tipping points 
- Danger of learning

- MAP

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.  [@Glaser2013](http://doi.org/10.1111/faf.12037 "Complex dynamics may limit prediction in marine fisheries")

Here we illustrate how a stochastic dynamic programming (SDP) algorithm can 
be driven by the predictions from a Gaussian Process (GP).
This provides two distinct advantages compared with contemporary approaches.
First, using a GP sidesteps the need for an accurate model-based description
of the system dynamics (or a model choice or model-averaging approach which
introduces yet more difficulty).  
Second, unlike parametric models which can only reflect 
uncertainty in the corresponding confidence intervals or posterior distributions
of their parameters, the GP approach provides a state-space dependent representation
of uncertainty.  This permits a much greater uncertainty far from the observed data than near the observed 
data.  These features allow the GP-SDP approach to find robust management solutions 
in face of both limited data and no knolwedge of correct model structure.

- Note on "not magic": honest uncertainty + SDP

The idea that any approach can perform well without either having to
know the model or have particularly good data should immediately draw
suspicion.  The reader must bear in mind that the strength of our approach
comes not from black-box predictive power from such limited information,
but rather, by providing a more honest expression of uncertainty outside
the observed data without sacrificing the predictive capacity near the
observed data. By coupling this more accurate description of what is known
and unknown to the decision-making under uncertainty framework provided
by stochastic dynamic programming, we are able to obtain more robust
management policies than with common parameteric modeling approaches.

- Note on comparing models (via value function rather than by "fit"). _Do we need this?_

- Note on terminology: "Non-parametric"  _Do we need this?_

## Background on the Gaussian Process

- Definition
- Previous application
- Why it is particularly suited to these two problems
- (Why this is a novel application thereof)

### Summary of approach

### Statement of the optimal control problem

- Underlying model 
- Available data 
- Value function

For simplicity we assume profit is simply linear in the realized harvest (only
enforcing the restriction that harvest can not exceed available stock)

### Parametric models

- Statement of the models

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model [Allen 2005](), the Myers
model. The familiar Ricker model involves two parameters, corresponding
to a growth rate and a carrying capacity, and cannot support alternative
stable state dynamics (though as growth rate increases it exhibits a
periodic attractor that proceeds through period-doubling into chaos. We
will generally focus on dynamics below the chaotic threshold for
the purposes of this analysis.) The Allen model resembles the Ricker
dynamics with an added Allee effect parameter [Courchamp](), below
which the population cannot persist.  The Myers model also has three
parameters and contains an Allee threshold, but has compensatory rather
than over-compensatory density dependence (resembling a Beverton-Holt
curve rather than a Ricker curve at high densities.)

We assume multiplicative log-normal noise perturbs the growth predicted 
by the each of the deterministic model skeletons described above. This 
introduces one additional parameter $\sigma$ that must be estimated by each
model.

<!-- equations just in appendix? -->

As we simulate training data from the Allen model (ref section), we will
refer to this as the structurally correct model.  The Ricker model is
thus a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure. Thus
we have three potential parametric models of the stock dynamics.

- Bayesian inference of parametric models

We infer posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (implemented in R [R]()
using jags, [jags]()).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures S1-S3, and Table S1, and the R code
provided). One-step-ahead predictions of these model fits are shown in
Figure 1.

- SDP via parametric models

An optimal policy function is then inferred through stochastic dynamic
programming for each model given the posterior distributions of the
parameter estimates.  This policy maximizes the expectation of the value
function integrated over the parameter uncertainty. (code implementing
this algorithm provided in the Appendix).

### The Gaussian Process model

- Statement of model

... more on GP ... [Munch 2005]()

We also estimate a simple Gaussian Process defined by
a radial basis function kernel of two parameters: $\ell$, which gives
the characteristic length-scale over which correlation between two 
points (e.g. any two points $X_t, X_{t+1}$, and $X_{t+\tau}, X_{t+1+\tau}$)
in state-space decays, and $\sigma$, which gives the scale of the 
process noise by which observations $Y_{t+1}$ may differ from their
predicted values $X_{t+1}$ given an observation of the previous state,
$X_t$.

- Inference of the model

Also unlike parametric models, this posterior distribution is still
conditional on the training data. As such, the uncertainty near the
observed data.

We use a Metropolis-Hastings Markov Chain Monte Carlo to infer posterior
distributions of the two parameters of the GP (Figure S4, code in
appendix), under weakly informative Gaussian priors (see parameters in
table S5). As the posterior distributions differ substantially from the
priors (Figure S4), we can be assured that most of the information in
the posterior comes from the data rather than the prior belief.

- SDP via the model

Though we are unaware of prior application of this type, it is reasonably
straight-forward to adapt the Gaussian Process for Stochastic Dynamic
Programming.  Recall that unlike the parametric models the Gaussian
process with fixed parameters already predicts a distribution of
curves rather than a single curve. We must first integrate over 
this distribution of curves given a sampling of parameter values drawn
from the posterior distribution of the two GP parameters, before
integrating over the posterior of those parameters themselves.

### Figure 1: Fitted Models

- All models fit the data quite well
- Information criteria would pick the simple, incorrect model.

(State-space figure??)

### Figure 2: Inferred Policies

- Inferred policies differ substantially among models
- The structurally correct model and the GP are close to the true model
- alternatives are not close

### Figure 3: Simulated results

-

### Figure 4: Robustness

- Results hold across range of parameters

(eek, distinguish better between result and discussion?)

- All models are "good fits" to the originally observed data. 
- (Simple model choice immediately leads us astray)

## Model definitions and estimation

Equation S1: Ricker model.

Figure S1: Ricker model: prior and posterior distributions for parameter estimates.

```{r, echo=FALSE, include=TRUE}
plot_ricker_traces
```

```{r, echo=FALSE, include=TRUE}
plot_ricker_posteriors
```
Table S1: Parameterization of the priors

```{r echo=FALSE, results="asis", include=TRUE}
xtable::xtable(data.frame(
  parameter = c("r0", "K", "sigma"),
  lower_bound = c(r0_prior_p[1], K_prior_p[1], stdQ_prior_p[1]),
  upper_bound = c(r0_prior_p[2], K_prior_p[2], stdQ_prior_p[2])))
```

Eq S2: Myers model 
Figure S2: Myers model: Traces, prior and posterior distributions for parameter estimates.

```{r, echo=FALSE, include=TRUE}
plot_myers_traces
```

```{r, echo=FALSE, include=TRUE}
plot_myers_posteriors
```

Table S2: Parameterization of the priors
```{r echo=FALSE, results="asis", include=TRUE}
xtable::xtable(data.frame(parameter = c("r0", "K", "theta", "sigma"),
           lower_bound = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           upper_bound = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2])))
```

Eq S3: Allen model

$$f(S_t) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - C\right)} $$

Figure S3: Allen model: prior and posterior distributions for parameter estimates.

```{r, echo=FALSE, include=TRUE}
plot_allen_traces
```

```{r, echo=FALSE, include=TRUE}
plot_allen_posteriors
```

Table S3: Parameterization of the priors

```{r echo=FALSE, results="asis", include=TRUE}
xtable::xtable(data.frame(parameter = c("r0", "K", "theta", "sigma"),
           lower_bound = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           upper_bound = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2])))
```

Eq S4: GP model 
Figure S4: GP model: prior and posterior distributions for parameter estimates.

```{r echo=FALSE, include=TRUE}
gp_assessment_plots
```

Table S4: Parameterization of the priors

## Optimal Control Problem

We seek the harvest policy $h(x)$ that maximizes:

$$ \max_{h_t} \sum_{t \in 0}^{\infty}  \Pi_t(X_t, h_t) \delta^t  $$

subject to the profit function $\Pi(X_t,h)$, discount rate $\delta$, and the state
equation

$$X_{t+1} = Z_t f(S_t)  $$
$$S_t = X_t - h_t $$

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape parameter
$\sigma_g$.

Form this we can write down the Bellman recursion as:

$$V_t(x_t) = \max_h \mathbf{E} \left(\Pi(h_t, x_t) + \delta V_{t+1}( Z_{t+1} f(x_t - h_t)) \right)$$

For simplicity we assume profit is simply linear in the realized harvest (only
enforcing the restriction that harvest can not exceed available stock), $\Pi(h,x) = min(h,x)$.

### Training data

Eacho of our models $f(S_t)$ must be estimated from training data, which
we simulate from the Allen model with parameters $r = $ ` r p[1]`, 
$K =$ ` r p[2]`, $C =$ ` r p[3]`, and  $\sigma_g =$ ` r sigma_g` 
for $T=$ `r Tobs$ timesteps, starting at initial condition $X_0 = $ `r Xo`.

-----------------------------------

<!-- OLD TEXT -->

@@

While methods that account for _parametric_ uncertainty can be very 
successful with the right model,
structural uncertainty of not knowing what model best approximates the 
dynamics poses considerably greater difficulty.

@@

@@

@@

<!-- More on the lack of data throughout the relevant state-space
     and how any concern about potential tipping points indicates 
     that the data exhibits this bias / problem.  
     -->

@@

@@

@@

<!-- 
### Quantitative vs Qualitative Decisions

@@

@@

In our example, we focus on the problem in which a manager must set
the harvest level for a marine fishery each year to maximize the net
present value of the resource, given an estimated stock size from the
year before.

### 2013-09-18 inst/doc/nonparametric-bayes.tex
writing writing

@@

We demonstrate how nonparametric Bayesian models can provide robust,
nearly-optimal solutions to decision making under uncertainty when
\emph{we don't know the correct model} to use.

While methods that account for \emph{parametric} uncertainty can be very
successful with the right model, structural uncertainty of not knowing
what model best approximates the dynamics poses considerably greater
difficulty.

Non-parametric Bayesian methods offer a promising statistical approach
for predictive modeling of ecological dynamics in regions of state space
where the data is adequate, while at the same time offering more
flexible patterns with greater uncertainty outside the observed data.
This contrasts from simple parametric models which provide relatively
constant level of uncertainty in regions with and without adequate data.
The consequence of such misplaced confidence outside the data can lead
to highly undesirable results that may be avoided with the more flexible
non-parametric Bayesian approach.

@@

Decision making under uncertainty is a ubiquitous challenge of natural
resource management and conservation.\\Here we illustrate how a
stochastic dynamic programming algorithm can be driven by the
predictions from a Gaussian process model, sidestepping the need for an
accurate model-based description of the system dynamics.

add references to (@Fischer2009, Polasky et al. 2011) as persective on
combining complex / critical dynamics + control theory

@@

Decision-theoretic (or optimal control) tools require a model that can
assign probabilities of future states (e.g.~stock size of a fishery)
given the current state and a proposed action (e.g.~fishing harvest or
effort). The decision maker then seeks to determining the course of
actions (also referred to as the policy) that maximizes the expected
value of some objective function (such as net present value derived from
the resource over time. (The approach can be adapted to permit
alternatives to maximizing expectation, such as minimizing the maximum
cost or damage that might be incurred. See Polasky et al. (2011)).
Management frequently faces a sequential decision-making problem --
after selecting an action, the decision-maker may receive new
information about the current state and must again choose an appropriate
action -- i.e.~setting the harvest limits each year based on stock
assessments the year prior.

More historical context, e.g. (Reed 1979, )

\paragraph{Parametric models}

Traditional approaches to optimal control (Pontryagin's principle,
stochastic dynamic programming) rely on knowledge of the state equation.
Such simple parametric models, frequently justified by mechanistic
underpinnings are used to provide probabilities of future states.
Fisheries literature, which enjoys some of the longest history in the
use of the optimal control framework, provides a clear example {[} {]} .
Models such as the Ricker or Beverton Holt curves are ubiquitous {[}too
strong? references?{]} Parametric models can reflect variability
introduced by a stochastic environment or demographic process my
incorporating random shocks from a given distribution.

@@

Unfortunately, estimates of parameter uncertainty are only as good as
the parametric models themselves. Too often, we do not understand the
system dynamics well enough to know if a model provides a good
approximation over the relevant range of states and timescales (criteria
that we loosely refer to as defining the ``right'' or ``true'' model.)
So called structural or model uncertainty is a more difficult problem
than parametric uncertainty. Typical solutions involve either model
choice, or model averaging. Either approach remains limited by the
degree to which any of the proposed models are sufficiently close to the
right model.

>>>>

Briefly set the stage: (see related papers)

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Decision-theory / optimal control framework
\item
  Fisheries context
\end{itemize}

@Fischer @Sethi2005, @Peretti, @Munch,

(Classical division)

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Dynamic uncertainty (stochasticity)
\item
  Parametric uncertainty
\item
  Structural uncertainty
\end{itemize}

\subsection{Two central challenges:}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  We don't know the model
\item
  We don't have data from where we need it most
\end{itemize}

\subsection{Why we don't know the model}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Complex dynamics
\item
  model choice and model averaging approaches
\end{itemize}

\subsection{Why we don't have data where we need it}

\begin{itemize}
\item
  Concerns of tipping points
\item
  Danger of learning
\item
  Note on ``not magic'': honest uncertainty + SDP
\item
  Note on comparing models (via value function rather than by ``fit'')
\item
  Note on terminology: ``Non-parametric''
\end{itemize}

\subsection{Background on the Gaussian Process}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Definition
\item
  Previous application
\item
  Why it is particularly suited to these two problems
\item
  (Why this is a novel application thereof)
\end{itemize}

\subsubsection{Summary of approach}

\subsubsection{Statement of the optimal control problem}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Underlying model
\item
  Available data
\item
  Value function
\end{itemize}

For simplicity we assume profit is simply linear in the realized harvest
(only enforcing the restriction that harvest can not exceed available
stock)

\subsubsection{Parametric models}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Statement of the models
\end{itemize}

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model \href{}{Allen 2005}, the
Myers model. The familiar Ricker model involves two parameters,
corresponding to a growth rate and a carrying capacity, and cannot
support alternative stable state dynamics (though as growth rate
increases it exhibits a periodic attractor that proceeds through
period-doubling into chaos. We will generally focus on dynamics below
the chaotic threshold for the purposes of this analysis.) The Allen
model resembles the Ricker dynamics with an added Allee effect parameter
\href{}{Courchamp}, below which the population cannot persist. The Myers
model also has three parameters and contains an Allee threshold, but has
compensatory rather than over-compensatory density dependence
(resembling a Beverton-Holt curve rather than a Ricker curve at high
densities.)

We assume multiplicative log-normal noise perturbs the growth predicted
by the each of the deterministic model skeletons described above. This
introduces one additional parameter $\sigma$ that must be estimated by
each model.

As we simulate training data from the Allen model (ref section), we will
refer to this as the structurally correct model. The Ricker model is
thus a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure. Thus
we have three potential parametric models of the stock dynamics.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Bayesian inference of parametric models
\end{itemize}

We infer posterior distributions for the parameters of each model in a
Bayesian context using Gibbs sampling (implemented in R \href{}{R} using
jags, \href{}{jags}). We choose uninformative uniform priors for all
parameters (See Appendix, Figures S1-S3, and Table S1, and the R code
provided). One-step-ahead predictions of these model fits are shown in
Figure 1.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  SDP via parametric models
\end{itemize}

An optimal policy function is then inferred through stochastic dynamic
programming for each model given the posterior distributions of the
parameter estimates. This policy maximizes the expectation of the value
function integrated over the parameter uncertainty. (code implementing
this algorithm provided in the Appendix).

\subsubsection{The Gaussian Process model}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Statement of model
\end{itemize}

\ldots{} more on GP \ldots{} \href{}{Munch 2005}

We also estimate a simple Gaussian Process defined by a radial basis
function kernel of two parameters: $\ell$, which gives the
characteristic length-scale over which correlation between two points
(e.g.~any two points $X_t, X_{t+1}$, and $X_{t+\tau}, X_{t+1+\tau}$) in
state-space decays, and $\sigma$, which gives the scale of the process
noise by which observations $Y_{t+1}$ may differ from their predicted
values $X_{t+1}$ given an observation of the previous state, $X_t$.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Inference of the model
\end{itemize}

Also unlike parametric models, this posterior distribution is still
conditional on the training data. As such, the uncertainty near the
observed data.

We use a Metropolis-Hastings Markov Chain Monte Carlo to infer posterior
distributions of the two parameters of the GP (Figure S4, code in
appendix), under weakly informative Gaussian priors (see parameters in
table S5). As the posterior distributions differ substantially from the
priors (Figure S4), we can be assured that most of the information in
the posterior comes from the data rather than the prior belief.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  SDP via the model
\end{itemize}

Though we are unaware of prior application of this type, it is
reasonably straight-forward to adapt the Gaussian Process for Stochastic
Dynamic Programming. Recall that unlike the parametric models the
Gaussian process with fixed parameters already predicts a distribution
of curves rather than a single curve. We must first integrate over this
distribution of curves given a sampling of parameter values drawn from
the posterior distribution of the two GP parameters, before integrating
over the posterior of those parameters themselves.

\subsubsection{Figure 1: Fitted Models}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  All models fit the data quite well
\item
  Information criteria would pick the simple, incorrect model.
\end{itemize}

(State-space figure??)

\subsubsection{Figure 2: Inferred Policies}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Inferred policies differ substantially among models
\item
  The structurally correct model and the GP are close to the true model
\item
  alternatives are not close
\end{itemize}

\subsubsection{Figure 3: Simulated results}

-

\subsubsection{Figure 4: Robustness}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Results hold across range of parameters
\end{itemize}

(eek, distinguish better between result and discussion?)

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  All models are ``good fits'' to the originally observed data.
\item
  (Simple model choice immediately leads us astray)
\end{itemize}

\subsection{Model definitions and estimation}

Equation S1: Ricker model.

Figure S1: Ricker model: prior and posterior distributions for parameter
estimates.

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-1.pdf}
\caption{plot of chunk unnamed-chunk-1}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-2.pdf}
\caption{plot of chunk unnamed-chunk-2}
\end{figure}

Table S1: Parameterization of the priors

\% latex table generated in R 3.0.1 by xtable 1.7-1 package \% Mon Sep 9
14:43:29 2013

\begin{table}[ht]
\centering
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{table}

Eq S2: Myers model Figure S2: Myers model: Traces, prior and posterior
distributions for parameter estimates.

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-4.pdf}
\caption{plot of chunk unnamed-chunk-4}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-5.pdf}
\caption{plot of chunk unnamed-chunk-5}
\end{figure}

Table S2: Parameterization of the priors \% latex table generated in R
3.0.1 by xtable 1.7-1 package \% Mon Sep 9 14:43:37 2013

\begin{table}[ht]
\centering
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & theta & 0.00 & 10.00 \\ 
  4 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{table}

Eq S3: Allen model

\[f(S_t) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - C\right)} \]

Figure S3: Allen model: prior and posterior distributions for parameter
estimates.

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-7.pdf}
\caption{plot of chunk unnamed-chunk-7}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-8.pdf}
\caption{plot of chunk unnamed-chunk-8}
\end{figure}

Table S3: Parameterization of the priors

\% latex table generated in R 3.0.1 by xtable 1.7-1 package \% Mon Sep 9
14:43:45 2013

\begin{table}[ht]
\centering
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & theta & 0.00 & 10.00 \\ 
  4 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{table}

Eq S4: GP model Figure S4: GP model: prior and posterior distributions
for parameter estimates.

\begin{verbatim}
$traces_plot
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-101.pdf}
\caption{plot of chunk unnamed-chunk-10}
\end{figure}

\begin{verbatim}

$posteriors_plot
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-unnamed-chunk-102.pdf}
\caption{plot of chunk unnamed-chunk-10}
\end{figure}

Table S4: Parameterization of the priors

\subsection{Optimal Control Problem}

We seek the harvest policy $h(x)$ that maximizes:

\[ \max_{h(t)} \sum_{t \in 0}^{\infty}  \Pi_t(X_t, h) \delta^t  \]

subject to the profit function $\Pi(X_t,h)$, discount rate $\delta$, and
the state equation

\[X_{t+1} = Z_t f(S_t)  \] \[S_t = X_t - h_t \]

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape
parameter $\sigma_g$.

Form this we can write down the Bellman recursion as:

\[V_t(x_t) = \max_h \mathbf{E} \left(\Pi(h_t, x_t) + \delta V_{t+1}( Z_{t+1} f(x_t - h_t)) \right)\]

For simplicity we assume profit is simply linear in the realized harvest
(only enforcing the restriction that harvest can not exceed available
stock), $\Pi(h,x) = min(h,x)$.

\subsubsection{Training data}

Eacho of our models $f(S_t)$ must be estimated from training data, which
we simulate from the Allen model with parameters $r = 2$, $K=8$, $C=5$
and $\sigma_g = 0.1$ for $T = 40$ timesteps, starting at initial
condition $X_0 = $.

\begin{center}\rule{3in}{0.4pt}\end{center}

@@

Further, the data on which a model has been parameterized frequently
fails to cover the possible state-space over which management decisions
must operate. Consequently a model do well in the region of state-space
in which it was estimated, but give erroneous confidence to predictions
outside of that region.

This problem is keenly felt in any system where a potential threshold or
tipping point is a concern. Such a tipping point, if it exists at all,
will lay outside the observed range of the observed data.

We demonstrate how nonparametric Bayesian models can provide robust,
solutions to decision making under uncertainty without knowing the
structural form of the true model.

@@

Decision making under uncertainty is a ubiquitous challenge of natural
resource management and conservation.\\Here we illustrate how a
stochastic dynamic programming (SDP) algorithm can be driven by the
predictions from a Gaussian Process (GP). This provides two distinct
advantages compared with contemporary approaches. First, using a GP
sidesteps the need for an accurate model-based description of the system
dynamics (or a model choice or model-averaging approach which introduces
yet more difficulty).\\Second, unlike parametric model-fitting
approaches, which can only reflect uncertainty in the corresponding
confidence intervals or posterior distributions of the parameters, the
GP approach provides a state-space dependent representation of
uncertainty. In other words, a GP estimated from a given data set
permits a much greater uncertainty far from the observed data than it
does near the observed data.

The idea that any approach can perform well without either having to
know the model or have particularly good data should immediately draw
suspicion. The reader must bear in mind that the strength of our
approach comes not from black-box predictive power from such limited
information, but rather, by providing a more honest expression of
uncertainty outside the observed data without sacrificing the predictive
capacity near the observed data.

By coupling this more accurate description of what is known and unknown
to the decision-making under uncertainty framework provided by
stochastic dynamic programming, we are able to obtain more robust
management policies than with common parameteric modeling approaches.

@@

Decision-theoretic or optimal control tools require a model that can
assign probabilities of future states (e.g.~stock size of a fishery)
given the current state and a proposed action (e.g.~fishing harvest or
effort).\\Management frequently faces a sequential decision-making
problem -- after selecting an action, the decision-maker may receive new
information about the current state and must again choose an appropriate
action -- such as setting the harvest limits each year based on stock
assessments the year prior.

The decision maker typically seeks to determining the course of actions
(also referred to as the policy) that maximizes the expected value of
some objective function such as net present value derived from the
resource over time.\\Though much can be said on how to choose this value
function appropriately (e.g.~see (Halpern et al.
2013)(http://doi.org/10.1073/pnas.1217689110 ``Achieving the triple
bottom line in the face of inherent trade-offs among social equity,
economic return, and conservation.'')) we will assume this is given.
(Nor is this approach necessarily constrained to maximizing the
expectated value of such a function - the decision-theoretic framework
can be adapted to alternatives such as minimizing the maximum cost or
damage that might be incurred; see Polasky et al. (2011)).

In representing future states with probabilities and maximizing
expectations, this approach provides a natural framework for handling
uncertainty.

The value function typically depends on the action or policy taken, as
well as the state of the system, in each interval of time. The state of
the system, in turn, is usually described by a dynamical model.

@@

Unfortunately, estimates of parameter uncertainty are only as good as
the parametric models themselves. Too often, we do not understand the
system dynamics well enough to know if a model provides a good
approximation over the relevant range of states and timescales (criteria
that we loosely refer to as defining the ``right'' or ``true'' model.)
So called structural or model uncertainty is a more difficult problem
than parametric uncertainty. Typical solutions involve either model
choice or model averaging. Either approach remains limited by the degree
to which any of the proposed models are sufficiently close to the right
model.
